Multi-level modeling with Openflexo/\FML - A contribution to the \mpc

### 2021-05-31 2-abstract.tex
Updates from Overleaf

\noteSylvain{Ce qui m'embête un peu avec cet abstract c'est qu'on met en avant la fédération de modèles alors que sur ce cas précis on en fait pas vraiment (sauf pour la partie outillage avec les éditeurs diagrammatiques). Concernant la fédération de modèle proprement dite, je crois qu'il faut qu'on montre qu'on a besoin d'une couche conceptuelle (qu'on relie à l'espace technique via les connecteurs techniques, les model slots et les rôles). C'est ce besoin qu'adresse le language \FML. On montre ici plutôt que le language \FML - utilisé comme language conceptuel - est assez intéressant, assez expressif pour capturer les exigences du challenge d'une part, et va jusqu'à permettre l'exécution des processus. Et cerise sur le gâteau: on peut construire un outillage graphique pour editer et exécuter ce use case.}
\noteSylvain{En revanche, on peut arguer que pouvoir au même niveau manipuler modèle et métamodèle nous simplifie grandement les choses pour adresser des problématiques multi-niveaux > un bout de la conclusion}.
%The models in a federation remain autonomous and represented in their original technological spaces %whereas virtual models and links serve as control components used to present different views to the %users and  maintain synchronization.

>>>>

%\noteSylvain{Ce qui m'embête un peu avec cet abstract c'est qu'on met en avant la fédération de modèles alors que sur ce cas précis on en fait pas vraiment (sauf pour la partie outillage avec les éditeurs diagrammatiques). Concernant la fédération de modèle proprement dite, je crois qu'il faut qu'on montre qu'on a besoin d'une couche conceptuelle (qu'on relie à l'espace technique via les connecteurs techniques, les model slots et les rôles). C'est ce besoin qu'adresse le language \FML. On montre ici plutôt que le language \FML - utilisé comme language conceptuel - est assez intéressant, assez expressif pour capturer les exigences du challenge d'une part, et va jusqu'à permettre l'exécution des processus. Et cerise sur le gâteau: on peut construire un outillage graphique pour editer et exécuter ce use case.}
%\noteSylvain{En revanche, on peut arguer que pouvoir au même niveau manipuler modèle et métamodèle nous simplifie grandement les choses pour adresser des problématiques multi-niveaux > un bout de la conclusion}.
%The models in a federation remain autonomous and represented in their original technological spaces %whereas virtual models and links serve as control components used to present different views to the %users and  maintain synchronization.

### 2021-05-31 6-modelpresentation.tex
Updates from Overleaf

In this section, we present the base metamodel presented in figure~\ref{fig:BaseMetamodel} with \textit{XSure} insurance domain use case, whose partial description was provided in the challenge description. The requirements \textbf{P1} to \textbf{P19} of XSure insurance domain are straightforwardly implemented by instantiating the \emph{XSure model}, as an instance of base metamodel (the left side of figure~\ref{fig:MultilevelArchitecture}).

@@

\todo{Gérer également la fin de P9, non traité pour le moment: artifact they produced must be associated with a validation task}

@@

The semantics is unclear relatively to the instantiation policy for artifacts. We assume that task execution implies that for each used and produced \emph{artifact type} defined in a related \emph{task type}, it exists at least one \emph{artifact} declaring required ontologic instantiation (through \textit{type} role). The following excerpt of \FML code shows a partial implementation of this. The same pattern applies to produced artifacts.
\noteFabien{Je ne suis pas bien sûr de comprendre le paragraphe}
\noteJC{autre remarque qui n'a rien à voir: si on suit les recommandations de l'appel, ce paragraphe devrait se retrouver dans la partie analyse étant donné qu'il établit un choix/une interprétation. Mais c'est délicat à intégrer sans le contexte du modèle de cette section. Ou alors il faudrait une liste, avec des références en avant}

@@

\begin{figure*}
 \centering
     \includegraphics[width=1.0 \textwidth]{Figures/AcmeFullArchitecture.pdf}
     \caption{Acme software development process architecture}
    \label{fig:AcmeFullArchitecture}
\end{figure*}

\todo[inline]{Augmenter la taille des fontes de la figure \ref{fig:AcmeFullArchitecture} <- a priori non}
\todo[inline]{Ajouter les programming languages <- je les vois apparaitre dans la figure, ou alors c'était un todo pour autre chose ?}

The Figure~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the full architecture. It also contains an instance of enacted a \textit{Software Engineering Process} at the bottom. The Acme metamodel specializes some concepts of the Base metamodel, for example, \texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType} extends \texttt{ArtifactType}. These concepts are further specialized  \texttt{CodingTaskType} extends \texttt{SETaskType} and \texttt{CodeArtifactType} extends \texttt{SEArtifactType}. 
The Acme metamodel also defines a specialized task type \texttt{Coding} extending \texttt{Task}, and provides the \texttt{Developer} concept extending \texttt{ActorType}. This metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the instances required to capture the challenge use case are defined in the final Acme model (lower part of the figure). Instances are represented with rounded boxes and linguistic instantiation is represented using dashed connectors.

@@

\textsf{'Requirements analysis'} is defined as an instance of \texttt{SETaskType}, \textsf{'Analyst'} is an instance of \texttt{ActorType} and \textsf{'Requirement specifications'} is an instance of \texttt{SEArtifactType}. The value of the property \texttt{producedArtifactTypes} of \textsf{'Requirements analysis'} is \texttt{\{}\textsf{Requirement specifications}\texttt{\}} and is \texttt{\{}\textsf{Analyst}\texttt{\}} for \texttt{allowedActorTypes} (\textbf{S1}). Similarly, \textsf{Test case design} is an instance of \texttt{SETaskType}, with a value of \texttt{true} for its property \texttt{isCritical}, and bound to \texttt{Analyst} by the \texttt{allowedActorTypes} relation. \textit{"Test case design"} produces \textsf{"Test cases"} defined as an instance of \textit{SETaskType}. The latter is used in "Test design review" (\textit{producedArtifactTypes} relation) (\textbf{S1} and \textbf{S13}, satisfied with \textbf{P9}). 
\noteFabien{pas de Test case design dans la figure}

The capture of \textbf{S3} requirement is performed while defining \textit{Coding} as an instance of concept \textit{CodingTaskType} extending \textit{SETaskType} and defining a relation \textit{languages} to \textit{ProgrammingLanguage} with 1..* cardinality. \textit{newTask(String)} behavior overrides generic implementation while instantiating a \textit{Coding} concept, specializing \textit{SEArtifact}, as shown in following excerpt of FML code:

@@

\textit{Code} concept extends \textit{SEArtifact}, and defines a relation \textit{language} to \textit{ProgrammingLanguage} with single cardinality (\textbf{S4}). \textbf{S5} is more ambigous as task type \textit{Coding} defines one or more programming languages but produces code which is expressed in one language only. This requirement is captured through the redefinition of \textit{declaresRequiredProducedArtifacts()} where programming language should also match. 
\textbf{S6} is guaranteed though the \textit{language} relation defined in \textit{Code} and following invariant declared in \textit{Code}:

\begin{lstlisting}
forEach (artifactType : types) {
  assert !(artifactType instanceof CodeArtifactType) | artifactType.doesImplement(languages)
}
\end{lstlisting}

\textit{Ann Smith} is the only one allowed to perform coding in COBOL (\textbf{S7}). This is implemented through the redefinition of \textit{newTask(String)} in \textit{CodingTaskType}:

@@

\textbf{S8} requirement is simply captured with the definition of \textit{Testing} instance of \textit{SETaskType}, \textit{Tester} instance of \textit{ActorType}, and \textit{Test report} instance of \textit{ArtifactType}.

\todo[inline]{requirement S9 non couvert pour le moment: compléter}

@@

\textbf{S13} requirement has been previously partially fulfilled. This must be completed with...

\todo[inline]{requirement S13 pas complètement couvert : compléter}

>>>>

In this section, we present the base metamodel presented in Figure~\ref{fig:BaseMetamodel} with \textit{XSure} insurance domain use case, whose partial description was provided in the challenge description. The requirements \textbf{P1} to \textbf{P19} of XSure insurance domain are straightforwardly implemented by instantiating the \emph{XSure model}, as an instance of base metamodel (the left side of figure~\ref{fig:MultilevelArchitecture}).

@@

@@

The semantics is unclear relatively to the instantiation policy for artifacts. In the requirements, nothing is explicitly stated about the link between the \emph{artifact type} of a \emph{task type} and the type of the corresponding artifact

We assume that task execution implies that for each used and produced \emph{artifact type} defined in a related \emph{task type}, it exists at least one \emph{artifact} declaring required ontologic instantiation (through \textit{type} role). The following excerpt of \FML code shows a partial implementation of this. The same pattern applies to produced artifacts.
\noteFabien{Je ne suis pas bien sûr de comprendre le paragraphe}

@@

\begin{figure*}
 \centering
     \includegraphics[width=1.0 \textwidth]{Figures/AcmeFullArchitecture.pdf}
     \caption{Acme software development process architecture (not all instances are shown for readability reasons)}
    \label{fig:AcmeFullArchitecture}
\end{figure*}

\todo[inline]{Ajouter les programming languages <- je les vois apparaitre dans la figure, ou alors c'était un todo pour autre chose ?}

The Figure~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the full architecture. This figure is only partial at instance level and not all instances are represented. It also contains an instance of enacted a \textit{Software Engineering Process} at the bottom. The Acme metamodel specializes some concepts of the Base metamodel, for example, \texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType} extends \texttt{ArtifactType}. These concepts are further specialized  \texttt{CodingTaskType} extends \texttt{SETaskType} and \texttt{CodeArtifactType} extends \texttt{SEArtifactType}. 
The Acme metamodel also defines a specialized task type \texttt{Coding} extending \texttt{Task}, and provides the \texttt{Developer} concept extending \texttt{ActorType}. This metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the instances required to capture the challenge use case are defined in the final Acme model (lower part of the figure). Instances are represented with rounded boxes and linguistic instantiation is represented using dashed connectors.

@@

\textsf{'Requirements analysis'} is an instance of \texttt{SETaskType}, \textsf{'Analyst'} is an instance of \texttt{ActorType} and \textsf{'Requirement specifications'} is an instance of \texttt{SEArtifactType}. The value of the property \texttt{producedArtifactTypes} of \textsf{'Requirements analysis'} is \texttt{\{}\textsf{Requirement specifications}\texttt{\}} and is \texttt{\{}\textsf{Analyst}\texttt{\}} for \texttt{allowedActorTypes} (\textbf{S1}). Similarly, \textsf{'Test case design'} is an instance of \texttt{SETaskType}, with a value of \texttt{true} for its property \texttt{isCritical}, and bound to \textsf{'Analyst'} by the \texttt{allowedActorTypes} relation. \textsf{'Test case design'} produces \textsf{'Test cases'} instance of \texttt{SETaskType}. The latter is used in \textsf{'Test design review'} (\texttt{producedArtifactTypes} relation) (\textbf{S1} and \textbf{S13}, satisfied with \textbf{P9}).

The capture of \textbf{S3} requirement is performed while defining \texttt{Coding} as an instance of concept \texttt{CodingTaskType} and defining a relation \texttt{languages} to \texttt{ProgrammingLanguage} with 1..* cardinality. The \texttt{newTask(String)} behavior overrides the generic implementation by instantiating a \texttt{Coding} concept, specializing \texttt{Task}, as shown in following excerpt of FML code:

@@

The \texttt{Code} concept extends \texttt{SEArtifact}, and defines a relation \texttt{language} to \textit{ProgrammingLanguage} with single cardinality (\textbf{S4}). 
\noteFabien{Code devient CodeArtifact}
\textbf{S5} is more ambiguous as a task type \texttt{Coding} defines one or more programming languages but produces code which is expressed in one language only. This requirement is captured through the redefinition of the behavior \texttt{declaresRequiredProducedArtifacts} where programming language should also match. 
\textbf{S6} is guaranteed though the \texttt{language} relation defined in \texttt{Code} and the following invariant declared in \texttt{Code}:  
\noteFabien{Coding devient CodingTask}
\begin{lstlisting}
forEach (artifactType : types) {
  assert !(artifactType instanceof CodeArtifactType) | artifactType.doesImplement(languages)
}
\end{lstlisting}

\textsf{Ann Smith} is the only one allowed to perform coding in COBOL (\textbf{S7}). This is implemented through the redefinition of \texttt{newTask(String)} in \texttt{CodingTaskType}:

@@

The requirement \textbf{S8} is simply expressed by the definition of the \textsf{'Testing'} instance of \texttt{SETaskType}, \textsf{'Tester'} instance of \texttt{ActorType}, and \textsf{'Test report'} instance of \texttt{ArtifactType}.

Critical task must additionnaly produce artifacts that must be associated with a validation task. This is modelled with a supplementary relation \texttt{validatedWith} defined in \texttt{Artifact} and referencing \texttt{Artifact} validating it (\textbf{S9}).

@@

\textbf{S13} requirement has been previously partially fulfilled. This must be completed with an association between artifact produced and the task that validates it. This is modelled by the reference to 'Test Cases' as a produced artifact type in 'Test Case design' and the reference to 'Test Cases' as a used artifact in 'Test Design review', as shown in figure \ref{fig:AcmeSoftwareDevelopmentProcess}.

### 2021-05-31 8-discussion.tex
Updates from Overleaf

  Due to the fact that our approach is level-agnostic, our solution could have had more
  or fewer levels depending on the variations of the use case. However, the number of levels is related
  to the problem, making the approach fixed. For example, given a requirement, an instance cannot be 
  changed into a concept easily.

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models.  Note that in our approach, the
  metamodels are built in a ad hoc way together with the models
  (co-construction), therefore the user also validates on a ad hoc basis.

@@

  Due to the level-agnostic aspect of our approach, we could insert a new
  level, for example. It would consist in creating new concepts we would link
  to other ones. Then, we would add any necessary constraints on those concepts
  and on the relationships linking them to the other existing concepts.

>>>>

  Due to the fact that our approach is level-agnostic, our solution could have had more
  or fewer levels depending on the variations of the use case. However, the number of levels is related
  to the problem, making the approach fixed.

@@

  Behaviors are continuously checked, ensuring the integrity of the models when
  changes occur. However, most of these behaviors have to be written by the
  users. Thus, the integrity of the contents relies on the users when they
  build the metamodels and the models.  Note that in our approach, the
  metamodels are built in an ad hoc way together with the models
  (co-construction), therefore the user also validates on a ad hoc basis.

@@

  Due to the level-agnostic aspect of our approach, we could insert a new
  level, for example. It would consist in creating new concepts we would link
  to other ones. Then, we would add any necessary constraints on those concepts
  and on the relationships linking them to the other existing concepts. This type of change can be easily done.

  On the other hand, making changes to the existing models could be difficult. For example, given a requirement, an instance cannot be changed into a concept easily. In the context of the process described in the challenge, this could

  Design Task ->

### 2021-05-31 9-relatedworks.tex
Updates from Overleaf

%Salvador commence
A plethora of multi-level modeling approaches and tools with different foundations have appeared in the recent years~\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

>>>>

%Salvador commence
A plethora of multi-level modeling approaches and tools with different foundations have appeared in the recent years\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.
